\section{Black-Scholes Equation}

\subsection{What is an Option?}

\begin{frame}{What is an Option?}
    \begin{itemize}
        \item Derivatives are financial contracts which derive their value from an underlying asset
        \item Futures, Forwards, Swaps, Options
        \item Options:
        \begin{itemize}
            \item Gives the right, not the obligation, to buy/sell asset at specified price on specified date
            \item Call gives right to buy
            \item Put gives right to sell
            \item American style can be utilised at any time up until expiration date
            \item European style can only be utilised on expiration date
        \end{itemize}
    \end{itemize}
\end{frame}

\subsection{Black-Scholes Equation}

\begin{frame}{Black-Scholes Background}
    \begin{itemize}
        \item Developed in 1970s by Fischer Black, Myron Scholes, and Robert Merton
        \item Transformation to the heat equation to find analytic solution to a second order parabolic PDE
        \item The solutions to this PDE are the options prices
        \item Leads to conclusion that options pricing dynamic is similar to heat dissipation
        \item Based on European style options
        \item American style options pricing can only be done numerically, not analytically
    \end{itemize}
\end{frame}

\begin{frame}{Black-Scholes Equation}
    The Black-Scholes equation is a second-order parabolic PDE. For a stock which does not pay dividends, we have:
    \[\frac{\partial c}{\partial t} + r S \frac{\partial c}{\partial S} + \frac{1}{2} \sigma^2 S^2 \frac{\partial^2 c}{\partial S^2} = rc\]
    \begin{columns}[T]
        \begin{column}{0.5\textwidth}
            \begin{itemize}
                \item $t$ = time
                \item $c$ = call option price
                \item $p$ = put option price
                \item $S$ = spot stock price
                \item $r$ = risk free interest rate
            \end{itemize}
        \end{column}
        \begin{column}{0.5\textwidth}
            \begin{itemize}
                \item $\sigma$ = stock volatility
                \item $S_0$ = spot stock price at time zero
                \item $K$ = strike price
                \item $T$ = time to maturity
            \end{itemize}
        \end{column}
    \end{columns}
    
    \vspace{10pt}
    Solutions to this PDE are the formulas for European call and put options prices.
\end{frame}

\begin{frame}{Boundary Conditions}
    \begin{itemize}
        \item For a European call option, we have the terminal condition:
        \[\text{payoff} = \max\left\{S_T - K,\, 0\right\}\]
        \item For a European put option, we have the terminal condition:
        \[\text{payoff} = \max\left\{K - S_T,\, 0\right\}\]
        where $t = T$ because this is the exact moment of expiration of the option
    \end{itemize}
\end{frame}

\subsection{Transformation to the Heat Equation}

\begin{frame}{Transformation to the Heat Equation}
    Firstly, we define:
    \[x = \ln\!\left(\frac{S(t)}{c(t)}\right) \qquad \text{and} \qquad \tau = \frac{1}{2} \sigma^2 (T - t)\]
    Take log of price as stock price follows geometric Brownian motion. \\
    Scale time by volatility. \\
    Define mapping from $c(s,t)$ to $f(x,\tau)$ with strike price $K$ as scale factor:
    \[c(s,t) = K f(x,\tau)\]
\end{frame}

\begin{frame}
    Substituting in the partial derivatives into Black-Scholes allows us to find the new PDE:
    \[\frac{\partial f}{\partial \tau} = \frac{\partial^2 f}{\partial x^2} + \left(\frac{2r}{\sigma^2} - 1\right) \frac{\partial f}{\partial x} - \frac{2r}{\sigma^2} f\]
    which can be re-written as:
    \[\frac{\partial f}{\partial \tau} = \frac{\partial^2 f}{\partial x^2} + (k - 1) \frac{\partial f}{\partial x} - k f\]
    where $k = \dfrac{2r}{\sigma^2}$. We see this is a PDE with similar form to the heat equation, but with extra terms $(k - 1)\,\dfrac{\partial f}{\partial x}$ and $-kf$.
\end{frame}

\begin{frame}
    To make the extra terms disappear, we define a new mapping:
    \[f(x,\tau) = \exp(\alpha x + \beta \tau)\, u(x,\tau)\]
    Drift is the expected return of an asset over time. We remove it and replace it with the risk-free rate to create a risk-neutral environment.
    The exponential function removes drift.
    We then solve and choose $\alpha$ and $\beta$ such that the extra terms are removed, so we find:
    \[\alpha = -\tfrac{1}{2}(k - 1) \qquad \text{and} \qquad \beta = -\tfrac{1}{4}(k + 1)^2\]
    which leads to a direct relationship between actual price of the option $c(s,t)$, and the modified price $u(x,\tau)$.
    Substituting in $\alpha$ and $\beta$ we find this relationship is given by:
    \[c(s,t) = K\, u(x,\tau)\, \exp\!\left(-\tfrac{1}{2}(k - 1)\,x - \tfrac{1}{4}(k + 1)^2\,\tau \right)\]
    which will later be used by the ANN.
\end{frame}

\begin{frame}
    We substitute this relationship into the original Black-Scholes equation, which cancels out the extra terms, so we are left with the heat equation, leading to the conclusion that options pricing dynamic is similar to heat dissipation. We have $x$ is the price of the stock after transformations, $\tau$ is the unit of time after transformations, and $u$ is the price of the option after transformations:
    \[\frac{\partial u}{\partial \tau} = \frac{\partial^2 u}{\partial x^2}, \qquad -\infty < x < \infty, \qquad \tau > 0\]
    Where we have the payoff boundary condition:
    \[u(x,0) = \max\!\left\{\exp\!\left(\tfrac{1}{2}(k+1)\,x\right) - \exp\!\left(\tfrac{1}{2}(k-1)\,x\right),\; 0\right\}\]
\end{frame}

\begin{frame}{Final Solution}
    We substitute the solution $u(x, \tau)$ of this heat equation into our expression:
    \[c(s,t) = K\, u(x,\tau)\, \exp\!\left(-\tfrac{1}{2}(k - 1)\,x - \tfrac{1}{4}(k + 1)^2\,\tau \right)\]
    We use the Gaussian kernel to solve this; we apply the solution formula, split the integral, and complete the square, allowing us to obtain the solutions to the Black-Scholes equation, which shows the pricing of European call and put options:
    \[c = S_0 N(d_1) - K\exp(-rT)\,N(d_2) \qquad p = K\exp(-rT)\,N(-d_2) - S_0 N(-d_1)\]
    where
    \[d_1 = \frac{\ln\!\left(\dfrac{S_0}{K}\right) + \left(r + \dfrac{\sigma^2}{2}\right)(T - t)}{\sigma\sqrt{T - t}} \qquad \text{and} \qquad d_2 = d_1 - \sigma\sqrt{T - t}\]
    $d_1$ and $d_2$ are parameters which represent risk-adjusted probabilities within the normal CDF $N(x)$, which comes from the Gaussian integral. This algebra follows from many lines of working whilst using the method of solving the heat equation, which we have learnt in this course.
\end{frame}
